---
title: "What CMRSS could delegate to `coin`, and what it could not"
subtitle: "Reuse, speed, and whether CMRSS and riposte should be one package"
author: "Jake Bowers and Claude"
date: 2026-09-07
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    embed-resources: true
  pdf:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
---

## The answer, before the argument

CMRSS does two separable things. It builds a reference distribution --- rank
scores, block-respecting draws of the treatment assignment, the exact
permutation mean and variance of each rank sum statistic, and the distribution
of the largest standardized one. And it computes an observed statistic ---
the smallest value the combined rank sum could take over the treated-unit
potential outcomes that the null hypothesis $H_{k,c}$ still permits, which in
the stratified case requires an integer program to allocate a budget of
large-effect units across blocks.

`coin::independence_test` computes the first of those. I checked its
permutation mean and standard deviation against the ones CMRSS computes and
they agree to six digits, and its Monte Carlo reference distribution matches
CMRSS's to Monte Carlo error. It computes none of the second, and neither does
SOUP, which was archived from CRAN on 2022-06-10 and whose menu of statistics
never accepted a user-supplied score vector anyway.

By executable lines of R, the reference-distribution layer is about a fifth of
the package. The rest stays. So the honest answer to "is this a lot of custom
code that duplicates a tested library" is: some of it is, and it is not where
the time goes.

The reference distribution is 95 percent of a single p-value, but the package
already lets you compute it once and pass it in, so that time is avoidable
without changing anything. The time in the confidence intervals goes somewhere
else --- into a function called `min_stat` --- and the fix for that is not
`coin` and not C++. It is the observation --- which I checked below
--- that `min_stat` is a step function of the threshold $c$ with a known,
finite set of jump points. That turns the confidence-interval search from an
approximate root-find into an exact search over a finite list.


## Why the package splits into layers at all

Before comparing CMRSS to anything, it helps to say what the package computes,
in the order it computes it, because the comparison has to go layer by
layer.

The hypothesis under test is $H_{k,c}$: at most $n - k$ of the units have an
individual treatment effect larger than $c$. Testing it needs three ingredients,
and turning the test into a confidence interval needs a fourth.

First, a test statistic. CMRSS uses a weighted stratified rank sum. Within
block $b$, each unit gets a score $a_b(j)$ depending only on its rank $j$ among
the $n_b$ outcomes in that block, and the block's contribution is the sum of
the scores of its treated units, divided by the number treated $m_b$ and
multiplied by a weight $w_b$. Summing across blocks gives one statistic. Using
several score functions --- Wilcoxon, Stephenson at several values of $s$,
polynomial at several powers $r$ --- gives $H$ statistics, and the package
combines them by standardizing each and taking the largest.

Second, a reference distribution for that combined statistic. The design fixes
$m_b$ treated units in each block, so re-drawing which units are treated,
independently across blocks, gives the randomization distribution. CMRSS draws
it by Monte Carlo.

Third --- and this is the ingredient with no counterpart in a general
permutation-testing package --- the value of the statistic to compare against
that distribution. Under a sharp null the researcher knows every unit's control
outcome, so the statistic is a number computed from the data. Under $H_{k,c}$
the researcher does not. The null permits many configurations of the control
potential outcomes, and each gives a different value of the statistic. The
p-value has to be valid for all of them, so CMRSS computes the *smallest*
value the statistic can take across the permitted configurations and compares
that. In one block, `min_stat` does this by imputing $Y_i - c$ for each treated
unit that the null holds to an effect of at most $c$, and $-\infty$ for the at
most $n - k$ units the null lets off, which pushes them to the bottom of the
ranking. Across blocks, the budget of exempt units has to be split among the
blocks in whatever way makes the combined statistic smallest, and that
allocation is the integer program in `R/solvers.R`.

Fourth, for the confidence intervals, the whole test gets inverted: for each
$k$, find the smallest $c$ at which the test no longer rejects.

Those four ingredients are the layers. `coin` owns the second and the exact
moments the second needs. It has no vocabulary for the third or the fourth.


## What `coin` reproduces

The statistic CMRSS builds is, in the terms `coin` uses, a linear statistic:
$T_h = \sum_i z_i g_h(i)$, where $z_i$ indicates treatment and $g_h(i)$ is a
score attached to unit $i$. CMRSS builds exactly this matrix of unit-level
scores inside `com_null_dist_block`, under the name `reordered_scores`, by
setting unit $i$'s entry in block $b$ to $w_b / m_b$ times its rank score.
Handing that matrix to `coin` should therefore give the same permutation mean
and covariance that `mu_sigma_list` computes in R.

It does, but only after fixing one thing that is easy to get wrong.

```{r}
#| label: moments
suppressMessages({library(coin); library(CMRSS)})
set.seed(20260907)

## Unbalanced blocks on purpose: 4 treated of 12. With m_b = n_b/2 the treated
## sum and the control sum have the same permutation distribution, so a
## balanced check cannot tell which of the two coin scored.
B <- 7; per <- 12; mtrt <- 4
n <- B * per
block <- factor(rep(seq_len(B), each = per))
Z <- unlist(lapply(split(seq_len(n), block),
                   function(ix) sample(c(rep(1, mtrt), rep(0, per - mtrt)))))

## Three polynomial score functions, one statistic each, same method in every block
mla <- lapply(c(2, 6, 10), function(r)
  lapply(seq_len(B), function(i)
    list(name = "Polynomial", r = r, std = TRUE, scale = FALSE)))

bs  <- CMRSS:::summary_block(Z, block)
w   <- CMRSS:::weight_scheme(bs, "asymp.opt")
sla <- lapply(mla, function(m) CMRSS:::score_all_blocks(bs$nb, m))

## The unit-level score matrix, built the way com_null_dist_block builds it
H <- length(mla)
reordered <- matrix(0, n, H)
for (h in 1:H) for (i in 1:B) {
  u <- bs$units.block[[i]]
  reordered[u, h] <- w[i] / bs$mb[i] * sla[[h]][[i]]
}
ms <- CMRSS:::mu_sigma_list(Z, block, w, mla, sla, bs)
```

With the treatment variable written as `factor(Z)`, `coin` takes the levels in
sorted order and scores the *first* one, which is the control group:

```{r}
#| label: trap
it0 <- independence_test(reordered ~ factor(Z) | block,
                         teststat = "maximum", alternative = "greater",
                         distribution = "asymptotic")
rbind(`coin linear statistic` = as.numeric(statistic(it0, "linear")),
      `sum over treated`      = colSums(reordered[Z == 1, ]),
      `sum over control`      = colSums(reordered[Z == 0, ]))
```

Reversing the levels puts `coin` on the treated units, and then its closed-form
moments and CMRSS's agree:

```{r}
#| label: agree
it1 <- independence_test(reordered ~ factor(Z, levels = c(1, 0)) | block,
                         teststat = "maximum", alternative = "greater",
                         distribution = "asymptotic")
rbind(`coin expectation` = as.numeric(expectation(it1)),
      `CMRSS mu`         = ms$mu,
      `coin sd`          = sqrt(diag(covariance(it1))),
      `CMRSS sigma`      = ms$sigma)
```

The level ordering changes which group is scored, it changes the answer, and it does not error.
Anyone who reuses `coin` this way needs a test that would catch it, which means
a test on an unbalanced design.

The reference distributions match too. Both draw within-block assignments with
the block's treated count fixed, so the only difference between them is Monte
Carlo noise. At 50,000 draws each:

```{r}
#| label: quantiles
#| eval: false
## continues the `moments` chunk above: same Z, block, mla, bs, w, sla, reordered
set.seed(11)
qs <- c(.5, .75, .9, .95, .975, .99)
nullC <- CMRSS:::com_null_dist_block(Z, block, mla, sla, null.max = 50000,
                                     weight = w, block.sum = bs)
it <- independence_test(reordered ~ factor(Z, levels = c(1, 0)) | block,
                        teststat = "maximum", alternative = "greater",
                        distribution = approximate(nresample = 50000))
round(rbind(CMRSS = quantile(nullC, qs), coin = qperm(it, qs)), 4)
```

```
         50%    75%    90%    95%  97.5%    99%
CMRSS 0.3129 0.9629 1.5418 1.9258 2.2468 2.5874
coin  0.3054 0.9629 1.5712 1.9258 2.2468 2.6242
```

Three of the six quantiles agree to the last digit printed, which happens
because the combined statistic on 7 blocks of 12 takes finitely many values and
both methods land on the same one. The largest of the three gaps that do appear
is 0.0368, the size of the Monte Carlo error at 50,000 draws, not a difference
in the statistic.

`coin` computes this distribution in C, through `libcoin`, and the gap widens
with $n$. Timings for the same three statistics at 10,000 draws, on this
machine:

```{r}
#| label: timing
#| eval: false
time_both <- function(B, per, nperm = 10000, seed = 1) {
  set.seed(seed)
  n <- B * per
  block <- factor(rep(seq_len(B), each = per))
  Z <- unlist(lapply(split(seq_len(n), block),
                     function(ix) sample(rep(0:1, each = per / 2))))
  mla <- lapply(c(2, 6, 10), function(r)
    lapply(seq_len(B), function(i)
      list(name = "Polynomial", r = r, std = TRUE, scale = FALSE)))
  bs  <- CMRSS:::summary_block(Z, block)
  w   <- CMRSS:::weight_scheme(bs, "asymp.opt")
  sla <- lapply(mla, function(m) CMRSS:::score_all_blocks(bs$nb, m))
  H <- length(mla); reordered <- matrix(0, n, H)
  for (h in 1:H) for (i in 1:B) {
    u <- bs$units.block[[i]]; reordered[u, h] <- w[i] / bs$mb[i] * sla[[h]][[i]]
  }
  t_cmrss <- system.time(CMRSS:::com_null_dist_block(
    Z, block, mla, sla, null.max = nperm, weight = w, block.sum = bs))[["elapsed"]]
  t_coin <- system.time(independence_test(
    reordered ~ factor(Z, levels = c(1, 0)) | block,
    teststat = "maximum", alternative = "greater",
    distribution = approximate(nresample = nperm)))[["elapsed"]]
  data.frame(blocks = B, n = n, CMRSS = t_cmrss, coin = t_coin,
             ratio = round(t_cmrss / t_coin, 1))
}
do.call(rbind, list(time_both(7, 12), time_both(20, 20), time_both(50, 20)))
```

```
 blocks    n CMRSS  coin ratio
      7   84 0.019 0.017   1.1
     20  400 0.694 0.096   7.2
     50 1000 1.665 0.120  13.9
```

CMRSS's time grows roughly with $n$ because it materializes an $n$-by-chunk
assignment matrix and multiplies; `coin`'s barely moves over this range.


## What `coin` does not reproduce

Four things. Three are structural and one is a design choice you could
revisit.

The attainable minimum under $H_{k,c}$ has no counterpart. `coin` computes the
statistic at the observed data. CMRSS needs the statistic minimized over the
control potential outcomes the null permits, which is `min_stat` in one block
and the integer program across blocks. Nothing in `coin`'s framework refers to
potential outcomes at all.

The confidence-interval inversion has no counterpart. `coin` has `confint`
methods for a few named location-shift tests; there is no general machinery
for inverting an arbitrary test over a threshold parameter.

The `comb.method = 2` reference distribution cannot be expressed as a `coin`
statistic. That path (`com_null_dist_block_stratum`) takes the maximum across
the $H$ methods *within* each block and then sums the block maxima. A maximum
of sums is a maximum of linear statistics, which is what `coin`'s
`teststat = "maximum"` computes; a sum of maxima is not linear in the
assignment vector, so `coin` cannot form it.

The completely randomized path combines by minimum p-value rather than by
maximum standardized statistic. `comb_null_dist_cre` converts each statistic to
its tail probability and takes the smallest, then calibrates that minimum
against its own randomization distribution. This is Tippett's combination, and
`coin` offers only `"maximum"` and `"quadratic"`. The two are close cousins ---
both are single-step methods calibrated on the joint null --- but they are not
the same function, and switching would change your operating characteristics.
This is the one item on this list that is a choice rather than a wall: the
draws underneath are still `coin`'s, and the min-p step on top is twelve lines.


## SOUP

SOUP implements Pesarin's nonparametric combination, which is the right family
of ideas --- Fisher, Liptak, Tippett, minP, maxT, direct sums --- and its
combining-function menu is broader than `coin`'s in exactly the direction the
CRE path needs. But it will not work here, for three reasons that compound.

CRAN archived it on 2022-06-10 for unresolved check failures, so it cannot be
a dependency of a package you intend to submit, and no one is maintaining it.

Its test statistics are a fixed list --- Anderson-Darling, difference of means,
$t$, Hotelling's $T^2$, and a linear-model coefficient. There is no argument
that accepts a user-supplied score vector, so Stephenson scores at $s = 30$
cannot be expressed in it.

And it has the same blind spot as `coin` about the third layer. Its stratified
analysis conditions on a categorical covariate, which handles the design, but
it computes the observed statistic from the observed data. The minimization
over null-permitted potential outcomes is simply not in the vocabulary.

I would close SOUP as an option and not revisit it.


## How much code this comes to

Counting executable lines --- neither blank nor comment --- in `R/`, and
assigning each function to exactly one layer:

| layer | lines | share |
|:---|---:|---:|
| Scores and design bookkeeping | 73 | 4% |
| Exact permutation moments | 94 | 5% |
| Drawing assignments, forming the reference distribution | 237 | 14% |
| The `comb.method = 2` statistic and reference distribution | 148 | 8% |
| The attainable minimum under $H_{k,c}$, and the solvers | 439 | 25% |
| p-value assembly and confidence-interval inversion | 491 | 28% |
| Benchmark wrappers around RIQITE | 264 | 15% |
| **total** | **1746** | |

The two layers `coin` computes are the second and third, 331 lines, 19 percent
of the package. Subtract the twelve lines of min-p calibration that `coin`
cannot do and it is 319 lines, 18 percent.

The `comb.method = 2` row absorbs `poly_sd` and `poly_strata_sd`, which return
$\sqrt{1/(2r-1) - 1/r^2}$. That is the standard deviation of $U^{r-1}$ for a
uniform $U$ --- the large-block limit of the polynomial score's spread --- not a
permutation moment of the realized design. `mu_sd_block` standardizes the
stratum-level statistics by it, so `coin` could not supply those numbers even
if the combination were expressible as one of its statistics.

So: about a fifth could be delegated. And that fifth contains almost none of
the code you would worry about, because it is the most mechanical and most
heavily exercised part of the package. The two rows that follow it in the
table --- the attainable minimum with its solvers, and the p-value assembly
with the interval inversion --- come to $439 + 491 = 930$ lines, and that is
where the mathematical content and the fragility both live.


## Where the time actually goes

Delegating a fifth of the lines only helps if it saves time. Here is what it
would save and what it would not.

For a single p-value at $n = 200$ across 10 blocks, three statistics, and
100,000 draws:

```{r}
#| label: profile-setup
#| eval: false
suppressMessages(library(CMRSS))
set.seed(99)
B <- 10; per <- 20; n <- B * per
block <- factor(rep(seq_len(B), each = per))
Z <- unlist(lapply(split(seq_len(n), block),
                   function(ix) sample(rep(0:1, each = per / 2))))
Y <- rnorm(n) + 0.5 * Z
mla <- lapply(c(2, 6, 10), function(r)
  lapply(seq_len(B), function(i)
    list(name = "Polynomial", r = r, std = TRUE, scale = FALSE)))
bs  <- CMRSS:::summary_block(Z, block)
w   <- CMRSS:::weight_scheme(bs, "asymp.opt")
sla <- lapply(mla, function(m) CMRSS:::score_all_blocks(bs$nb, m))
ms  <- CMRSS:::mu_sigma_list(Z, block, w, mla, sla, bs)
m <- sum(Z); k <- floor(0.9 * m)
```

```{r}
#| label: profile-pval
#| eval: false
t_null <- system.time(CMRSS:::com_null_dist_block(
  Z, block, mla, sla, null.max = 1e5, weight = w, block.sum = bs))[["elapsed"]]
t_coef <- system.time(cf <- CMRSS:::comb_matrix_block(
  Z, Y, block, 0, mla, sla, bs))[["elapsed"]]
t_lp   <- system.time(CMRSS:::solve_optimization(
  Z, block, w, cf, m - k, ms, TRUE, bs, "highs"))[["elapsed"]]
c(null = t_null, coef = t_coef, ilp = t_lp)
```

```
  null distribution   3.432 s
  coefficient matrix  0.031 s
  integer program     0.159 s
```

Those three add to 3.622 seconds, of which the reference distribution is 3.432,
or 95 percent. For a single p-value, then, `coin` would cut the wall clock by
about a factor of ten.

Before reaching for `coin`, though: `pval_comb_block` already takes a
`stat.null` argument. The reference distribution depends on the design and the
score functions, not on the outcomes, so across simulation replicates with a
fixed design it can be computed once and passed in. That is free, needs no new
dependency, and saves the same 3.4 seconds. In this project's
`Analysis/sim_test_fns.R` the question is moot for a different reason: the
Stephenson comparison there already calls `coin::independence_test` directly,
with a note recording that `CMRSS::pval_comb_block` was dropped because
$H_{k,c}$ is the wrong hypothesis for a sharp-null power comparison. So the
simulations never run this code.

The confidence intervals are where the time goes, and there the picture
reverses. At the same design, with only 1,000 draws and a tolerance of 0.05,
one call to `com_block_conf_quant_larger(set = "treat")` took 21.8 seconds. I
counted the calls and profiled the time:

```{r}
#| label: profile-ci
#| eval: false
## continues the `profile-setup` chunk above
cnt <- new.env(); cnt$lp <- 0L; cnt$cf <- 0L
trace(CMRSS:::solve_optimization, tracer = quote(cnt$lp <- cnt$lp + 1L), print = FALSE)
trace(CMRSS:::comb_matrix_block,  tracer = quote(cnt$cf <- cnt$cf + 1L), print = FALSE)
Rprof("ci.out", interval = 0.01)
ci <- com_block_conf_quant_larger(Z, Y, block, set = "treat",
        methods.list.all = mla, opt.method = "ILP_highs",
        null.max = 1000, tol = 0.05, alpha = 0.1)
Rprof(NULL)
untrace(CMRSS:::solve_optimization); untrace(CMRSS:::comb_matrix_block)
c(lp = cnt$lp, cf = cnt$cf)
head(summaryRprof("ci.out")$by.self, 6)
```

```
solve_optimization calls: 520    comb_matrix_block calls: 520
total time: 21.84 s

               self.time self.pct total.time total.pct
"solver_run"        7.71    35.30       7.71     35.30
"match.arg"         2.67    12.23       6.26     28.66
"order"             2.17     9.94       5.82     26.65
"sort.list"         1.05     4.81       7.80     35.71
"rank"              0.92     4.21      10.39     47.57
"min_stat"          0.41     1.88      13.42     61.45
```

Read the `total.pct` column. `min_stat` accounts for 61 percent of the run, and
`rank` --- called from inside it --- for 48 percent. HiGHS itself is 35 percent,
and `match.arg`, which is argument checking inside the `highs` package's R
interface rather than any solving, is another 12.

Each call to
`comb_matrix_block` loops over $H = 3$ methods, $B = 10$ blocks, and
$m_b + 1 = 11$ values of the exempt count, calling `min_stat` once each:
$3 \times 10 \times 11 = 330$ calls. The confidence-interval search made 520
such calls, so `min_stat` ran $330 \times 520 = 171{,}600$ times, and each run
does a full `rank()` over its block.

So the reference distribution, which is the only thing `coin` would replace, is
not on this list at all. Delegating it changes 21.8 seconds into 21.8 seconds.


## The change I would actually make

`min_stat` is called 171,600 times at what are mostly repeated values of $c$,
and it does not need to be. As a function of $c$ with everything else held
fixed, it is a step function, it is monotone, and its jump points are known in
advance.

Here is why. Inside `min_stat`, the treated units the null exempts are chosen
by their outcome rank, through `sort_treat(Y, Z)`, which looks at $Y$ and not
at $c$. So the exempt set does not move as $c$ changes. What $c$ changes is the
imputed value $Y_i - c$ for the non-exempt treated units, and the statistic
depends on those only through their ranks among all $n_b$ values in the block.
A treated unit's rank relative to another treated unit never changes, because
both shift by the same $c$. A treated unit $i$ crosses a control unit $j$
exactly when $c$ passes $Y_i - Y_j$. So the statistic is constant between
consecutive values of the set $\{Y_i - Y_j : Z_i = 1, Z_j = 0\}$, which has at
most $m_b (n_b - m_b)$ elements.

Checked on a 12-unit block with 6 treated, where that bound is $6 \times 6 = 36$:

```{r}
#| label: stepfun
set.seed(7)
nb <- 12; mb <- 6
Zb <- sample(rep(0:1, each = mb))
Yb <- round(rnorm(nb), 3)
score <- CMRSS:::rank_score(nb, list(name = "Polynomial", r = 6,
                                     std = TRUE, scale = FALSE))

diffs <- sort(unique(outer(Yb[Zb == 1], Yb[Zb == 0], "-")))
grid  <- seq(min(diffs) - 1, max(diffs) + 1, length.out = 20001)
step  <- diff(grid)[1]

for (k in c(nb, nb - 2, nb - 5)) {
  vals  <- vapply(grid, function(cc)
    CMRSS:::min_stat(Zb, Yb, k, cc, score = score), numeric(1))
  jumps <- grid[which(diff(vals) != 0) + 1]
  cat(sprintf(
    "k = %2d: %3d distinct values, %2d jumps; every jump at a Y1-Y0 difference: %s; monotone: %s\n",
    k, length(unique(vals)), length(jumps),
    all(vapply(jumps, function(x) min(abs(x - diffs)) <= step, logical(1))),
    all(diff(vals) <= 0)))
}
cat(sprintf("candidate breakpoints m_b(n_b - m_b) = %d; distinct differences = %d\n",
            mb * (nb - mb), length(diffs)))
```

Two consequences follow, and the second matters more than the first.

The coefficient matrices can be built once instead of 520 times. For each
method $h$, block $b$, and exempt count, sort that block's at most
$m_b(n_b - m_b)$ differences and evaluate `min_stat` at each, giving a lookup
table. Every later value of $c$ becomes a `findInterval` call. In the profiled
design that replaces 171,600 evaluations with $3 \times 10 \times 11 \times 100
= 33{,}000$ done once. I have not built this, so I will not quote a speedup; the
61 percent is measured and the replacement is projected.

More important, the confidence limits stop being approximate. The integer
program's objective depends on $c$ only through those step functions, so it is
itself a step function of $c$, with jumps confined to the union of the blocks'
breakpoints --- at most $\sum_b m_b (n_b - m_b)$ values, which is 1,000 in the
profiled design. The exact confidence limit is one of those values. A bisection
over the sorted candidate list finds it in about $\log_2(1000) \approx 10$
solver calls per $k$, and the bracket shrinks as $k$ decreases because the code
already sets `c.max <- quantiles[k + 1]`.

That would let you delete this, which appears four times in the two
confidence-interval functions:

```r
c.sol <- uniroot(f, interval = c(c.min, c.max), extendInt = "downX", tol = tol)$root
c.sol <- round(c.sol, digits = -log10(tol))
if (f(c.sol) <= 0) {
  while (f(c.sol) <= 0) { c.sol <- c.sol - tol }
  c.sol <- c.sol + tol
} else {
  while (f(c.sol) > 0) { c.sol <- c.sol + tol }
}
```

Those `while` loops step by `tol` from a rounded root with no iteration cap,
and each step calls the solver again. They exist because `uniroot` assumes a
continuous $f$ and $f$ is a step function, so `uniroot` returns a point that
need not satisfy the inequality the code wants. Searching the breakpoints
directly removes the assumption violation rather than patching around it, and
returns the exact limit instead of one accurate to `tol`.

One caution. `min_stat` breaks rank ties with `ties.method = "first"`, so its
value *at* a breakpoint depends on that convention. An exact search should
evaluate just below and just above each candidate rather than at it, and the
tests should pin down which side the reported limit takes.

A second, independent change sits next to this one, and it takes less work. `match.arg` at 12 percent of self
time is argument checking inside `highs`, repeated on all 520 calls for a
problem with $\sum_b (m_b + 1) + H + 1 = 114$ variables. A 114-variable linear
program does not take 15 milliseconds to solve; it takes that long to set up
through the R interface. Only the objective coefficients change between calls,
so building the model once and updating those would cut into the 35 percent as
well.


## On moving to C++

I would not, and the profile is the reason.

HiGHS is already compiled C++; `coin` and `libcoin` are already compiled C. The
two places where CMRSS's own R code is hot are the reference distribution ---
which `coin` computes in C, and which the existing `stat.null` argument lets
you compute once anyway --- and `min_stat`, whose 61 percent disappears under
the precomputation above, because the work is redundant rather than slow. After
those two changes there is no R-level hot spot left for C++ to attack. Writing
one would add a compilation toolchain requirement to a package whose remaining
bottleneck is 520 round trips into someone else's compiled solver.

If a C++ target ever does appear, the honest place to look for it first is
riposte's `perm.R`, whose own header calls it "the seam for a future
C++/Rust backend" and which is the only loop over permutations in that
package.


## Should CMRSS and riposte be one package

No, and the reason is the null hypothesis rather than any judgment about code
overlap.

The two packages share four of the five things I listed at the start. Both
build a unit-level score matrix from a transformation of the outcome. Both need
the exact permutation moments of a linear statistic under block randomization
--- CMRSS in `mu_sigma_list`, riposte in `riposte_sw_moments`, and I verified
above that `coin` reproduces CMRSS's, while riposte's function documentation
records that it was verified against `coin` too. Both draw block-respecting
assignments. Both combine several statistics into one and refer the result to
the randomization distribution.

They differ at the third layer, and the difference is not bridgeable by a
shared function. riposte tests the sharp null that every individual effect is
zero, so it knows every unit's control outcome and evaluates its statistic at
the observed data --- `crossprod(draws, scores)`, row one. CMRSS tests
$H_{k,c}$, which leaves the control outcomes partly unknown, so its observed
statistic is the solution of an optimization problem. A package built around
the first cannot express the second by adding an argument.

They also differ in how they combine. riposte uses a quadratic form with the
full permutation covariance, a Cauchy combination of mid-p values, and a
maximum. CMRSS uses a maximum of standardized statistics in the stratified case
and a minimum p-value in the completely randomized case. There is no conflict
here --- these are five combining functions and both packages could in
principle offer all five --- but it does mean the combination layer is not a
single shared object today.

And the authorship differs. CMRSS is Kim and Li's method with you as a
contributor; riposte is yours. Merging would tangle the citation and the
maintenance.

What I would share instead is the *test contract*, which riposte has already
adopted: `coin` in `Suggests`, and tests that check the permutation moments and
the reference-distribution quantiles against it. CMRSS does not currently have
those tests. Adding them takes little work and no runtime dependency, and it is
the answer to "is our custom code as well tested as the library's" --- because
it makes the library the test. The unbalanced-block check in this memo is the
version of that test I would write first, since the balanced version would have
passed while `coin` scored the wrong group.


## Two smaller things I noticed

`pval_comb_block` returns `mean(stat.null >= stat.min)`, which does not include
the observed assignment in either the numerator or the denominator. riposte's
`riposte_perm_pvalue` does include it, giving
$(1 + \#\{\text{draws} \ge \text{observed}\}) / (1 + R)$, where $R$ is the
number of draws. The gap is of order $1/R$, so at the default 100,000 draws it
is $10^{-5}$, far below anything that
changes a decision. I flag it as a convention difference between your two
packages rather than as a defect. Under $H_{k,c}$ the quantity
being compared is an optimization minimum rather than the statistic at the
observed assignment, so whether the usual $+1$ argument transfers is a question
for the paper's validity theorem, not one I can settle from the code.

`coin` will give the maximum statistic's reference distribution asymptotically,
through a multivariate normal on the standardized statistics, which would make
the confidence-interval threshold `thres` deterministic and remove Monte Carlo
error from the endpoints. I mention it because it is available and takes one line, not
because I recommend it: Kim and Li's validity is a finite-sample randomization
argument, and trading it for an asymptotic one to save a threshold calculation
is a bad exchange.


## What I would do, in order

1. Add tests to CMRSS that check `mu_sigma_list` and the reference-distribution
   quantiles against `coin`, on an unbalanced design, with `coin` in
   `Suggests`. This is the smallest change on the list and it answers the
   testing question directly.

2. Replace `min_stat`'s repeated evaluation with a per-block step-function
   table, and replace the `uniroot`-plus-`while` search with a bisection over
   the union of breakpoints. This gives exact confidence limits, deletes the
   four uncapped `while` loops, and removes the 61 percent.

3. Leave the reference distribution in R, but make sure every caller that can
   reuse `stat.null` does. If profiling after step 2 still shows it, swap in
   `coin` behind the existing function signature for the `Z.perm = NULL` path,
   which is the one `pval_comb_block` uses. `coin` has no way to accept a
   supplied assignment matrix, so the `Z.perm` path that
   `com_block_conf_quant_larger` passes through would have to stay in R.

4. Do not write C++. Do not use SOUP. Keep CMRSS and riposte separate.
